Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-09-20 manuscript/main.tex
Updates from Overleaf
Added parametiersation section

\documentclass[12pt, a4paper]{article}
\usepackage{graphicx} % Required for inserting images
\usepackage[parfill]{parskip} % Removes indent from new paragraphs
\usepackage{amsmath, amssymb, amsfonts}
\usepackage[colorlinks = true,
			citecolor  = blue]{hyperref}

@@

\begin{abstract}
    The objective of this research is to compare the computational methods used to estimate Bayesian stochastic volatility models. A simulation study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo and their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these sampling strategies are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which algorithm is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy.
\end{abstract}

@@

\section{Background and Literature Review}
    Stochastic volatility models are used in financial econometrics to model the variance or volatility of returns in a financial instrument. The return of a financial asset often exhibits patterns in the variance which can be modelled to improve prediction and estimates of risk or tail events. Unlike other volatility models such as ARCH or GARCH, stochastic volatility models treat the variance as a random variable. Furthermore, these models have a state space representation since the variance is expressed as a latent stochastic process. State space models can be difficult to estimate since there is a latent state for every data observation in the time series, which means there are often more parameters than data points. These models can be estimated using frequent frameworks but this research will focus on Bayesian estimation strategies.

@@

    \subsection{Estimation strategy}
        Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.} \citep{dejong1995}.

@@

    \subsection{Design}
        Simulation Based Calibration (SBC) is a technique that checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples conditional on the simulated data.

@@

        ESS measures the efficiency of the MCMC sampler. It calculates the number of (effectively) independent draws from the posterior draws generated by a Markov chain. A poor ESS can arise from high autocorrelation in the Markov chain which leads to highly dependent samples. An efficient MCMC algorithm takes less resources (for example, time and number of draws) to get a representative sample of the target distribution. If a MCMC algorithm possesses higher ESS for the majority of its parameters (relative to another strategy), then we may conclude that this method is a more efficient sampler (conditional on the model).

@@

@@

>>>>

\documentclass[12pt, a4paper]{article}
\setcounter{tocdepth}{2}
\usepackage{graphicx} % Required for inserting images
\usepackage[parfill]{parskip} % Removes indent from new paragraphs
\usepackage{amsmath, amssymb, amsfonts}
\usepackage[colorlinks = true,
			citecolor  = blue]{hyperref}

@@

\begin{abstract}.  
    The objective of this research is to compare the computational methods used to estimate Bayesian stochastic volatility models. A simulation study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo and their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these sampling strategies are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which algorithm is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy.
\end{abstract}

\newpage

\tableofcontents{\protect\newpage}

@@

\section{Background and Literature Review}

    \subsection{Stochastic Volatility}
        Stochastic volatility models are used in financial econometrics to model the variance or volatility of returns in a financial instrument. The return of a financial asset often exhibits patterns in the variance which can be modelled to improve prediction and estimates of risk or tail events. Unlike other volatility models such as ARCH or GARCH, stochastic volatility models treat the variance as a random variable. Furthermore, these models have a state space representation since the variance is expressed as a latent stochastic process. State space models can be difficult to estimate since there is a latent state for every data observation in the time series, which means there are often more parameters than data points. These models can be estimated using frequent frameworks but this research will focus on Bayesian estimation strategies.

@@

    \subsection{Gaussian Mixture Approximation}
        Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.} \citep{dejong1995}.

@@

    \subsection{Simulation Design}
        Simulation Based Calibration (SBC) is a technique that checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples conditional on the simulated data.

@@

        ESS measures the efficiency of the MCMC sampler. It calculates the number of effectively independent draws from the posterior draws generated by a Markov chain. A poor ESS can arise from high autocorrelation in the Markov chain which leads to highly dependent samples. An efficient MCMC algorithm takes less resources (for example, time and number of draws) to get a representative sample of the target distribution. If a MCMC algorithm possesses higher ESS for the majority of its parameters (relative to another strategy), then we may conclude that this method is a more efficient sampler (conditional on the model).

@@

    \subsection{Model Parameterisation}

        The parameterisation of a model can affect the performance of a MCMC algorithm when sampling from models with complex posterior geometries. An example of this is Neal's funnel \citep{neal2003slice} where the Hamiltonian Monte Carlo sampler encounters performance issues in hierarchical models and produces biased samples.

        The stochastic volatility model as described in section 2.1 is follows a "centered" parameterisation. This describes the central location of the latent state parameters which are centered around the mean and lag of the log volatility $\mu +\phi(h_t - \mu)$. A model can be reparameterised such that the states are sampled on a distribution centered on 0 (i.e "non centered") which are later transformed to have the correct mean.  It is useful to compare model parameterisations since sampling strategies may be sensitive to the posterior geometry.

        \subsubsection{Non centered Hamiltonian Monte Carlo}
        First sample from a standard normal distribuition and multiply by the variance of the log volatility.

        $$
        \begin{aligned}
        h_{std} \sim& \space normal(0,1) \\
        h =& h_{std} \times \sigma_{\eta} \\ 
        h \sim& normal(0, \sigma_{\eta}^2)
        \end{aligned}
        $$

        The states are sampled with the mean centered at zero. Then apply the appropriate rescaling to get samples from the log volatility.

        $$
        \begin{aligned}
        h_1 =& \space \frac{h_{std, 1}\times \sigma_{\eta}} {\sqrt{1 - \phi^2}} + \mu \\
        h_{t+1} =& \space h_{std, t+1}\times \sigma_{\eta} + \mu  + \phi(h_{t} - \mu),\space t\neq 1
        \end{aligned}
        $$

        This returns log volatility as desired.

        $$
        \begin{aligned}
        h_1 \sim& \space normal \left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
        h_{t+1} \sim& \space normal(\mu +\phi(h_t - \mu) , \sigma_{\eta}^2), \space\space t\neq 1\\ 
        \end{aligned}
        $$

        \subsubsection{Non centered Gaussian Mixture}
        The non centered model for the Gaussian approximation is expressed slightly differently due to the use of the Kalman Filter. While HMC can sample from the joint posterior directly with state vectors centered at zero, applying the Kalman Filter requires the state equation to be a function of the state variable.

        Starting with the log chi squared model:

        $$
        \begin{aligned}
        y^{\ast}_t =& h_t + z_t \\
        h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t
        \end{aligned}
        $$

        Expand the state equation:

        $$
        \begin{aligned}
        h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
        =& \space \mu + \phi h_t - \phi\mu + \sigma_{\eta} \eta_t  \\
        =& \space \mu (1- \phi) + \phi h_t + \sigma_{\eta} \eta_t  \\
        \end{aligned}
        $$

        Take out the intercept and put it in the measurement equation:

        $$
        \begin{aligned}
        y^{\ast}_t =& h_t + \mu (1- \phi) + z_t \\
        h^{\ast}_{t+1} =& \space \phi h_t + \sigma_{\eta} \eta_t  \\
        \end{aligned}
        $$

        The mean of the log volatility is now inside the measurement equation and de-meaned from the state equation.

        \subsubsection{Prior Predictive check}
        It is worth noting that the centered and non-centered parameteristions are the same stochastic volatility model. Different parameteristions express the same mathematical models in ways that may make it easier or harder for a given algorithm to sample.

        To demonstrate this, a prior predictive simulation of the second state parameter is performed for both a centered and non centered model (using the paramterisation described in section 3.3.1). This is the same as the first two steps of SBC, except instead of generating a dataset, generate the parameters implied by the prior draw. Let $\boldsymbol{\theta}$ be a vector of static parameters.

        $$
        \begin{aligned}
        \boldsymbol{\theta}^{sim} \sim \pi(\boldsymbol{\theta})
        \end{aligned}
        $$

        Generate a the second state parameter implied by the joint prior.

        $$
        \begin{aligned}
        h_2^{sim} \sim \pi (h_2|h_1, \boldsymbol{\theta}^{sim})
        \end{aligned}
        $$

        Generating 1000 samples of $h_2^{sim}$ from both centered and non-centered models gives samples from the same data generating process as shown in figure x.

        % \begin{figure}
        %     \centering
        %     \includegraphics[scale=0.1]{ppc_h2.png}
        %     \caption{Prior predictive samples of second state variable}
        % \end{figure}

        % Emphasise here that different paramterisations of the models are the same model
        % Use a prior predictive check from both paramterisations show that the results are
        % the same

\section{Results}

    \subsection{Hamiltonian Monte Carlo}

    \subsection{Gaussian Mixture Approximation}

    \subsection{Reweigthing rank statistics of Gaussian Mixture}

@@

\newpage
